---
title: "**Variables and Measures of Cognition**"
author: "Gabriel I. Cook"
#date: "`r Sys.Date()`"
date: "`r format(Sys.time(), '%d %B, %Y')`"

execute:
  #enabled: false
  freeze: auto
---

```{r message=FALSE, warning=FALSE, include=FALSE}
# secret functions
# source(here::here("r", "src", "color_format_text.R"))
```


```{r}
#| label: load-packages
#| include: false

#library()
#source(here::here("r", "my_functions.R"))
source(here::here("r", "_common.R"))
R.utils::sourceDirectory(here::here("r", "functions"))
```

# **Overview**

In this module, we will consider different cognitive tasks, understand how those are measured, and understand the variables associated with them.   

## **Readings and Preparation**

There are no formal readings assigned for this topic. We will use class time to discuss the symmetry span and the go/no-go tasks, how they are performed by participants, what variables are measured, and how those tasks are correlated with other variables. 

Because you will read papers related to cognitive tasks in order to inform your team project, paper that introduce either complex-working memory span tasks (e.g., symmetry span) or the go/no-go task would benefit you during class time. Reading through a relevant paper would help you answer questions and conceptualize the task and its relation to the larger project.  


```{r eval=FALSE, warning=FALSE, include=FALSE}
# **To Do**
# 
## **Supplementary Readings**

#- [R Workflow Basics](https://r4ds.hadley.nz/data-transform)
```
