---
title: "*Legends and arrangement*"
author: "Gabriel I. Cook"
#date: "`r Sys.Date()`"
date: "`r format(Sys.time(), '%d %B, %Y')`"

execute:
  #enabled: false
  freeze: auto

#order: 2 
---

```{r}
#| label: load-packages
#| include: false
```


```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# to restart a fresh session
.rs.restartR()
```


::: callout-important
## Under construction. 

This page is a work in progress and may contain areas that need more detail or that required syntactical, grammatical, and typographical changes. If you find some part requiring some editing, please let me know so I can fix it for you.

:::


```{r include=FALSE}
#base_plot +
#  guides(col = guide_legend(
#    title.theme = element_text(
#      size = 15,
#      face = "italic",
#      colour = "red",
#      angle = 0
#    )
#  )
#  )
```

# **Overview**

Out of the box, **{ggplot2}** creates some really good data visualizations. When you map variables to aesthetics other than `x` or `y`, the plot will include a legend. A plot legend represents a key labeling system for understanding different data by colors, patterns, and symbols. For example, a legend could contain a variable name, a listing of variable levels, and some color or shape aesthetic used to visually discriminate those variable levels. The legend labels assumes some position, some font, some format, and some color. Aesthetic properties of a legend take on some color, shape, size, and orientation. In most cases, these elements match the aesthetic properties of the plot but they do not need to and they can be customized to suit the needs of a given plot. In fact, all properties of a plot legend can changed in **{ggplot2}**. 

If a legend is created automatically with some `geom_*()`, it can be removed. If a legend is not created automatically, it can be added to the plot using functions like `ggplot2::legend()`. And importantly, legends can be modified to suit out needs as describe in this module. A basic reason to change the legend appearance without changing the plot is to make the legend more readable and user friendly.

# **To Do**

## **Readings**

Reading should take place in two parts:

 - *Prior to class*, the goal should be to familiarize yourself and bring questions to class. The readings from [TFDV](https://clauswilke.com/dataviz) are conceptual and should facilitate readings from [EGDA](https://ggplot2-book.org/) for code implementation.
 - *After class*, the goal of reading should be to understand and implement code functions as well as support your understanding and help your troubleshooting of problems.

*Before Class*: First, read to familiarize yourself with the concepts rather than master them. Understand why one would want to visualize data in a particular way and also understand some of the functionality of **{ggplot2}**. I will assume that you attend class with some level of basic understanding of concepts.

*Class*: In class, some functions and concepts will be introduced and we will practice implementing **{ggplot2}** code. On occasion, there will be an assessment involving code identification, correction, explanation, etc. of concepts addressed in previous modules and perhaps some conceptual elements from this week's readings. 

*After Class*: After having some hands-on experience with coding in class, homework assignments will involve writing your own code to address some problem. These problems will be more complex, will involving problem solving, and may be open ended. This is where the second pass at reading with come in for you to reference when writing your code. The module content presented below is designed to offer you some assistance working through various coding problems but may not always suffice as a replacement for the readings from **Wickham, Navarro, & Pedersen (under revision)**. *ggplot2: Elegant Graphics for Data Analysis (3e)*.

- [Wilke (2019). *Fundamentals of Data Visualization*. Redundant coding](https://clauswilke.com/dataviz/redundant-coding.html)


## **External Functions**

Provided in class:

`view_html()`: for viewing data frames in html format, from `/r/my_functions.R` 

You can use this in your own work space but I am having a challenge rendering this of the website, so I'll default to `print()` on occasion.

```{r}
source(here::here("r", "my_functions.R"))
```

## **Libraries**

- **{dplyr}** `r packageVersion("dplyr")`: for selecting, filtering, and mutating
- **{magrittr}** `r packageVersion("magrittr")`: for code clarity and piping data frame objects
- **{ggplot2}** `r packageVersion("ggplot2")`: for plotting
- **{forcats}** `r packageVersion("forcats")`: for factor reordering


## **Load libraries**

```{r message=FALSE}
library(dplyr)
library(magrittr)
library(ggplot2)
library(forcats)
```


# **Loading Data**

For this exercise, we will use some data from a 2023 CMS swim meet located at: "https://github.com/slicesofdata/dataviz23/raw/main/data/swim/cleaned-2023-CMS-Invite.csv".

```{r}
SWIM <- readr::read_csv("https://github.com/slicesofdata/dataviz23/raw/main/data/swim/cleaned-2023-CMS-Invite.csv",
                        show_col_types = F)
```


# **Legend-Related Functions**

## `ggplot2::guide_legend()` 

`guide_legend()` is the go to function for customizing both the appearance and behavior of legends. For example, if the legends require some fine-tuning from their out-of-the-box behavior, you can override its behavior. The goal should be to improve the overall presentation in order to facilitate perception, interpretation, memory for, and decision-making about the data visualization representing the data.  

There is much you can do with `guide_legend()`:

```
guide_legend(
  title = waiver(),
  title.position = NULL,
  title.theme = NULL,
  title.hjust = NULL,
  title.vjust = NULL,
  label = TRUE,
  label.position = NULL,
  label.theme = NULL,
  label.hjust = NULL,
  label.vjust = NULL,
  keywidth = NULL,
  keyheight = NULL,
  direction = NULL,
  default.unit = "line",
  override.aes = list(),
  nrow = NULL,
  ncol = NULL,
  byrow = FALSE,
  reverse = FALSE,
  order = 0,
  ...
)

```

## `ggplot2::guides()`

`guides()` can be used to set guides for each scale. The documentation explains also that the guides can be set for each scale individually using the `guide` argument, or en masse with `guides()`. For example, `scale_shape(guide = "legend")` will 


Because the legend is also part of an overall plot theme, some legend adjustments can be made using `ggplot2::theme()`.

## `scale_<aesthetic>_<type>()` functions

Guides can be specified in each `scale_<aesthetic>_<type>()` function  or more generally in `guides()`.  As the **{ggplot}** documentation explains, `guide = "legend"` used inside a `scale_<aesthetic>_<type>()` , function is "*syntactic sugar*" for `guide = guide_legend()`. For example, `scale_color_manual(guide = "legend")` would achieve the same outcome as `guides(col = guide_legend())`. A benefit for the latter is that you may not need to remember the aesthetic and type of the variable mapped. In order to learn more about setting specific characteristics of each scale in more detail, review the documentation for `guides()`.


# **Creating Some Base Plots**

We are going to create some plot objects that we will modify later to illustrate how to use the functions mentioned above to change characteristics of legends in our data visualizations.  

**Plot A**: A plot for which we map a single variable to the `col` aesthetic:

```{r}
(base_plot <- SWIM |>
   filter(Event == "Freestyle") |>
   filter(Team != "Mixed") |>
   filter(Distance == 100) |>
   ggplot(mapping = aes(x = Split50, 
                        y = Time,
                        col = Team
                        )
          ) +
   geom_point(position = position_jitter(), alpha = .7) 
)

#base_plot + labs(title = "", tag = "Base Plot")
```

**Plot B**: For another plot, we map multiple variables to different aesthetics. We will map a single variable to the `col` aesthetic and a single variable to the `size` aesthetic. In this case, `col` is mapped to a discrete variable and `size` is mapped to a continuous variable. 

```{r}
(base_plot_2 <- base_plot +
   geom_point(mapping = aes(col = Team, 
                            size = Time
                            ),
              position = position_jitter(), 
              alpha = .7
              ) 
)
```

**Plot C**: We can also create a plot that maps a single variable to multiple aesthetics. We will map a single variable to the `size` aesthetic and a single variable to both the `col` aesthetic and the `shape` aesthetic. When groups are encoded using two aesthetics rather an a single one, this is referred to as *redundant encoding*. 

Claus Wilke discusses Redundant encoding in a    [chapter](https://clauswilke.com/dataviz/redundant-coding.html) of his textbook and we will use redundant encoding to address plot limitations associated with users who have color-vision deficiencies or simply to make the variable levels more distinctive from one another. Redundant encoding is a way to make data visualizations more perceptually efficient  which we will discuss in a later module. 

In this plot, you will see that the legend also encoded redundantly such that the levels of team vary both in color and in shape. Because team is also confounded with time, the different shapes also differ in size although this is because of a different dimension.

```{r}
(base_plot_3 <- base_plot +
   geom_point(mapping = aes(size = Time,
                            shape = Team,
                            fill = Team
                            ),
              position = position_jitter(), 
              alpha = .7,
              col = "grey20",
              stroke = 1
              ) +
   scale_shape_manual(values = c(21, 24))    # filled circles and triangles
)
```

Along with graduate advisor [Steve Franconeri](https://www.kellogg.northwestern.edu/faculty/directory/franconeri_steven.aspx), [Christine Nothelfer](www.cnothelfer.com) and colleagues have studied this how redundant encoding helps with visual selection processes like segmentation and grouping. A brief overview can be found on [Nothelfer's website](https://www.cnothelfer.com/redundant-encoding) and a related publication is [Nothelfer et al., (2017). Redundant encoding strengthens segmentation and grouping in visual displays of data](https://www.cnothelfer.com/_files/ugd/f70594_6e1ad80beb2340cb9e3f96982140f590.pdf). 

**Plot D**: We can also create a plot for which we map the same variable to different aesthetics. Here, `Team` to `col`, `size`, and `shape`. Because all of the aesthetics are mapped to the same variable, there is only *one legend* that encodes all three aesthetics. 

```{r}
(base_plot_4 <- base_plot +
   geom_point(mapping = aes(col = Team, 
                            size = Team,
                            shape = Team,
                            alpha = Team
                            ),
              position = position_jitter()
              )
)
```

Note: The variables are mapped to aesthetics to illustrate certain functionality rather than appropriateness. Remember that certain aesthetics are designed for specific types of variables, for example, discrete or continuous. If you match variables to aesthetics that violate these expectations you will receive warnings as shown below. 

```
Warning messages:
1: Using size for a discrete variable is not advised. 
2: Using alpha for a discrete variable is not advised. 
```


## **Examining Base Plots: Legend Elements**

In the initial base plot, you can see that the `Team` variable mapped to the `col` aesthetic appears in the legend positioned to the right of the plot. There is a title, which inherently takes on the name of the column variable in the data frame. There are keys, which inherently take on the values of the variations (e.g., levels). If the variable mapped to the aesthetic is a constant, or has no variation or levels within its vector, the legend will nevertheless appear but will present only a single key. In such instances, a legend likely has little to no perceptual utility and should either be removed from the plot and/or be set manually using one of the `scale_<aesthetic>_manual()` functions. 

In the additional base plots, legends again appear to the right of the plot. There are titles for each legend as well as their keys. When there is more than one legend, legends are ordered positionally. 


# **Changing Legend Spatial Positioning**

By default, legends will appear to the *right* of the plot. Legend position, however, can be adjusted to reflect changes in location like left, top, bottom, or location as specified by xy coordinates in the plot space, coordinates that appear inside the plot rather than outside the plot. One of the easiest ways to change the legend position is by using the `theme()` function and by setting an argument value for `legend.position`, including `"none"`, `"left"`, `"right"`, `"bottom"`, `"top"`, or a two-element numeric vector containing x and y coordinate values.

## **Removing a Legend**

The most crude way to change a legend position is to remove it completely from the visualization. Although the plots have more than one key in their legend, which add some perceptual utility, there may be instances where you would wish to remove the plot completely. For instance, perhaps you use direct labeling, annotation, or some other detail that obviate the legend's utility. 


### Using `theme()` with a single legend:

```{r}
base_plot + theme(legend.position = "none")
```

### Using `theme()` with multiple legends:

When there is more than one legend, all legends will be removed when set to `legend.position = "none"`. 

```{r}
base_plot_2 + theme(legend.position = "none")
```


### Using `guides()` with a single legend:

`guides(<aesthetic> = "none")`

The `guides()` function allows you to change many legend properties. Although the syntax is a little bit more complicated, `guides()` along with helper function `guide_legend()` used to control the legend guide may provide greater flexibility in the long run.

Because the plot contains a variable mapping to `col`, the legend can also be removed using  `guides()` and either set the aesthetic element to `"none"` or `FALSE`. As you will see with other functions, however, `FALSE` may be deprecated. In addition, although in many examples you will see the aesthetic referenced by its full name `color`, using its abbreviated name `col` will achieve the same outcome. For this purpose, my examples in this module will use the abbreviated form so that it matches that which I use in `mapping = aes()`. 

```{r}
base_plot + guides(col = "none")

# or base_plot + guides(col = FALSE)

# or base_plot + guides(color = "none") will also work
```

### Using `guides()` with a multiple legends:

With Plot 2, there is both `col` and `size`, so we would specify one or both.

Remove the color legend:

```{r}
base_plot_2 + guides(col = "none")

# or base_plot_2 + guides(col = FALSE)
```

Remove the size legend:

```{r}
base_plot_2 + guides(size = "none")
```

Remove both legends:

```{r}
base_plot_2 + guides(col = "none", 
                     size = FALSE
                     )
```


### *Using `scale_<aesthetic>_<type>()` with a single legend*:

When using `scale_*()` functions, you need to set `guide = "none"` as the use of `FALSE` has been deprecated.

Remove the color legend:

```{r}
base_plot +
  scale_color_discrete(guide = "none")
```

Remove the size legend:

```{r}
base_plot +
  scale_size(guide = "none")
```

Remove both legends:

```{r}
base_plot_2 +
  scale_color_discrete(guide = "none") +
  scale_size(guide = "none")
```

When using specific `scale_<aesthetic>_<type>()` functions, however, you must ensure that they are applied according to their aesthetic and type or that functions assuming a particular aesthetic and type (e.g., `scale_color_gradient()`) adhere to the aesthetic and type defined in the plot object. For example, because the variable type is discrete, `scale_color_manual(guide = "legend")`, `scale_color_continuous(guide = "legend")`,  `scale_color_binned(guide = "legend")` and some others will throw errors although `scale_color_hue(guide = "legend")`, `scale_color_brewer(guide = "legend")` will not throw errors. You just need to remember that your functions need to match the aesthetic and type already used in the plot object.
this


## **Repositioning Legends (Right, Left, Top, Bottom)**

The default position is `"right"`. Changing the spatial positioning of the legend can be achieved using the same `theme()` function and by setting the `legend.position` argument to `"left"`, `"top"`, `"bottom"`. Only some of these position modifications will be illustrated here. You can also achieve this using the `guides()` and `guide_legend()` combination illustrated earlier.

### Using `theme()` with a single legend:

```{r}
base_plot +
  theme(legend.position = "top")
```

### Using `theme()` with multiple legends:

```{r}
base_plot_2 +
  theme(legend.position = "top")
```



## **Repositioning Legends (Changing their Spatial Order)**

When you have more than one legend, their ordering can be rearranged using `guides()` and by specifying an `order` within helper function `guide_legend()`. To control each legend specifically, remember the guide argument is the aesthetic itself as seen here.

`guides(<aesthetic> = guide_legend())`

```
guides(
  col = guide_legend(),
  fill = guide_legend(),
  shape = guide_legend(),
  size = guide_legend()
  )
```

## **Creating a Complex Plot**

We will create a more complex plot to  better illustrate reordering methods. 

```{r}
(plot_complex <- SWIM |>
  filter(Event %in% c("Breaststroke", "Backstroke")) |>
  filter(Distance <= 200) |>
  mutate(Distance = factor(Distance)) |>
  ggplot(mapping = aes(x = Split50, 
                       y = Time,
                       fill = Team,
                       size = Time,
                       col = Event,
                       shape = Distance
                       )
         ) +
  geom_point()
)
```

`plot_complex` contains *four* legends. The variables are ordered from top to bottom: `Time`, `Event`, `Distance`, and `Team` and the aesthetics are ordered: `size`, `col`, `shape` and `fill`. These orders indicated that legends do not appear arranged alphabetically by variable name or aesthetic. Rather than worry about how they are ordered by default, lets just concern ourselves with arranging the order to what we want. 

```{r}
plot_complex

plot_complex +
  guides(
    col = guide_legend(order = 1),
    fill = guide_legend(order = 3),
    shape = guide_legend(order = 2),
    size = guide_legend(order = 4)
  )
```

The numbers do not need to be sequential but rather just differ in magnitude.

```{r}
plot_complex +
  guides(
    col = guide_legend(order = 21),
    fill = guide_legend(order = 1),
    shape = guide_legend(order = 38),
    size = guide_legend(order = 49)
  )
```


Note: If you wish to make changes to a legend corresponding to a continuous aesthetic like  `col` (or `color`) may be, `guide_legend()` will not work. You will need to use `guide_colorbar()` as show here.

```{r}
SWIM |>
  filter(Event %in% c("Breaststroke", "Backstroke")) |>
  filter(Distance <= 200) |>
  mutate(Distance = factor(Distance)) |>
  ggplot(mapping = aes(x = Split50, 
                       y = Time,
                       fill = Team,
                       size = Time,
                       col = Time,
                       shape = Distance
                       )
         ) +
  geom_point() +
  guides(
    col = guide_colorbar(order = 1),
    fill = guide_legend(order = 2),
    shape = guide_legend(order = 3),
    size = guide_legend(order = 4)
  )
```


## **Repositioning Legends (xy Coordinates)**

In addition to global positioning, you can have more direct control over the exact coordinates of the legend position if the gross location options are not appropriate. This type of repositioning is necessary when you want to position a legend in the site the plot itself rather than next to it. 

The fine-grained tuning is achieved by specifying a two-element vector for the xy coordinates of the plot but not according to the x and y axis scales. Using four xy coordinate pairs, we can see that the plot ranged from 0,0 to 1,1 so our numeric values need to fall between 0 and 1 inclusive. 

```{r}
suppressWarnings(plot(gridExtra::arrangeGrob(
  base_plot + 
    theme(legend.position = c(0, 0)) + 
    labs(title = "legend.position = c(0, 0)"),
  
  base_plot + 
    theme(legend.position = c(0, 1)) + 
    labs(title = "legend.position = c(0, 1)"),
  
  base_plot + 
    theme(legend.position = c(1, 0)) + 
    labs(title = "legend.position = c(1, 0)"),
    
  base_plot + 
    theme(legend.position = c(1, 1)) + 
    labs(title = "legend.position = c(1, 1)"),
  ncol = 2
  ))
  )
```

By deduction, `legend.position = c(.5, .5)` will position the legend in the plot center.

```{r}
base_plot + 
    theme(legend.position = c(.5, .5))
```

We can also place it more strategically someplace in the bottom right. 

```{r}
base_plot + 
    theme(legend.position = c(.8, .2))
```

Note: Depending on the plot dimensions and the uniformity of the x and y axis scales, you may need to experiment a bit.   


# **Legend Keys and Labels**

## **Adjusting Legend Label Position**

The text labels corresponding to the aesthetics can be rearranged by setting `label.position` to `"right"`, `"left"`, `"top"`, `"bottom"` (e.g., `guide_legend(label.position = "top")`). You can also remove the labels using `guides(col = guide_legend(label = FALSE))` although a color or shape seen in a plot without a corresponding label would be confusing. 

### **Removing a Legend Labels**

`guides(<aesthetic> = guide_legend(label = FALSE))`

```{r}
base_plot_2 +
  guides(col = guide_legend(label = FALSE))
```

But what do the colors represent?


### **Changing Legend Label Position**

```{r}
suppressWarnings(
  plot(gridExtra::arrangeGrob(
    base_plot_2 + 
      guides(col = guide_legend(label.position = "right")) + # default
      labs(title = 'label.position = "right"', tag = "A"),
    
    base_plot_2 + 
      guides(col = guide_legend(label.position = "left")) +
      labs(title = 'label.position = "left"', tag = "B"),
    
    base_plot_2 + 
      guides(col = guide_legend(label.position = "top")) +
      labs(title = 'label.position = "top"', tag = "C"),
    
    base_plot_2 + 
      guides(col = guide_legend(label.position = "bottom")) +
      labs(title = 'label.position = "bottom"', tag = "D"),
    ncol = 2
  ))
) 
```

### **Changing the Labels Direction/Orientation**

Legend labels are often presented vertically when legends are placed to the right or left of the plot and presented horizontally when oriented to the top or bottom of the plot. You may with to change this detail.

#### Adjusting generally using `theme()`

`theme(legend.direction = "")` will adjust all legends to be `"horizontal"` or `"vertical"` (default).

```{r}
base_plot_2 + 
  theme(legend.direction = "vertical")

base_plot_2 + 
  theme(legend.direction = "horizontal")
```


#### Adjusting specifically using `guides()` and `guide_legend()

`guides(<aesthetic> = guide_legend(direction = ""))`

```{r}
base_plot_2 + 
  guides(col = guide_legend(direction = "vertical"),
         size = guide_legend(direction = "horizontal")
         )

base_plot_2 + 
  guides(col = guide_legend(direction = "horizontal"),
         size = guide_legend(direction = "horizontal")
         )
```

#### Adjusting the orientation and the location with `theme()` and `guides()`:

```{r}
base_plot_2 + 
  theme(legend.direction = "horizontal", 
        legend.position = "bottom"
        )


base_plot_2 + 
  guides(col = guide_legend(direction = "horizontal"),
         size = guide_legend(direction = "horizontal")
         ) +
  theme(legend.position = "top")
```


# **Finer Tuning of Legends**

You can change other characteristics of legends using the `theme()` function as shown below. There our additional characteristics that you can change with other functions. 

```
legend.background,
  legend.margin,
  legend.spacing,
  legend.spacing.x,
  legend.spacing.y,
  legend.key,
  legend.key.size,
  legend.key.height,
  legend.key.width,
  legend.text,
  legend.text.align,
  legend.title,
  legend.title.align,
  legend.position,
  legend.direction,
  legend.justification,
  legend.box,
  legend.box.just,
  legend.box.margin,
  legend.box.background,
  legend.box.spacing,
  panel.background,
```

# **Changing Legend Point Size using `guides()`**

As you have seed, when you add aesthetics to plots, the plot will contain a legend providing a reference key for that aesthetic. You may have  noticed that the shapes in the legend can often appear quite small and are often difficult to process. We have addressed how to change point size in `geom_point()` by either setting or mapping a variable in the data frame to the `size` aesthetic. A cognitive limitation associated with small-sized shapes is placed on the user because those smaller shapes demand more effect to perceive and interpret. When the user needs to distinguish between the colors of small shapes, the smaller the shapes are, the more difficult that cognitive becomes. Moreover, distinguishing between two smaller shapes is more difficult than distinguishing between larger shapes. When a variable is mapped to `size`, however, some of the points become larger in size, making them more easy to process but there may be times you would just like to make the shapes a little more prominent in the legend.

The `guides()` function is describes in the docs as *"Guides for each scale can be set scale-by-scale with the guide argument, or en masse with guides()."*

## **Comparing Legend Point Size in Plots**

Let's create some data visualizations in order to investigate the legend properties. One plot will reflect the default behavior of `geom_point()` adding a legend to a plot corresponding to the mapping a variable to the color aesthetic using `col = Team`. Another plot will map `col = Team` but also set `size = 4`. A final plot will map `col = Team` and map `size = Team` so the size will be determined by the geom.

```{r}
plot1 <- SWIM %>%
  filter(Team != "Mixed",
         Time < 500
         ) %>%
  ggplot(data = ., 
         mapping = aes(x = Time, y = Split50)) +
  geom_point(aes(col = Team)) +
  labs(title = "default",
       tag = "A"
       )

plot2 <- SWIM %>%
  filter(Team != "Mixed",
         Time < 500
         ) %>%
  ggplot(data = ., 
         mapping = aes(x = Time, y = Split50)) +
  geom_point(size = 4, aes(col = Team)) +
  labs(title = "size = 4",
       tag = "B"
       )

plot3 <- SWIM %>%
  filter(Team != "Mixed",
         Time < 500
         ) %>%
  ggplot(data = ., 
         mapping = aes(x = Time, y = Split50)) +
  geom_point(aes(size = Team, col = Team)) + # note the warning: Using size for a discrete variable is not advised.
  labs(title = "aes(size = Team)",
       tag = "C"
       )

suppressMessages(
  plot(
    gridExtra::arrangeGrob(plot1, plot2, plot3, ncol = 1)
  )
  
)
```

All three plots contain legends but they differ in the rendering of point size. Plot A has the smallest circles and Plot C has variation in circle size because both `col` and `size` are mapped to the variable. Importantly, the point size in the legend and in the plot are the same and this characteristic is important mapping a variable to `size`. Mismatching point sizes between the plot and the legend would certainly be confusing. When size is a constant, however, changing their size in the legend can ease the processing demand. The point size on the legend may result in difficulty with seeing the color, even for those which normal color vision. You don't want your audience to squint when you are given your talk or stare at the legend in an attempt to understand the color differences.  


## **Adjusting Keys in Legends**

For various reasons, you may need to adjust the orientation, size, color, or some aesthetic property of legend keys in order to make plots more user friendly. We will work through some examples of these modifications using `guide_legend()` for a given aesthetic.

### **Reversing the Legend Keys**

If your legend order can be reversed to solve a perceptual inconsistency, just reverse them. Reversing the order may be a solution to some problems and works easily when there are only two groups but such a simple fix may not work when there are three or more groups to label.

`guides(<aesthetic> = guide_legend(reverse = TRUE))`

```{r}
suppressMessages(
  plot(
    gridExtra::arrangeGrob(plot1, 
                           
                           plot1 + labs(title = 'guides(col = guide_legend(reverse = TRUE))') + 
                             guides(col = guide_legend(reverse = TRUE)), 
                           ncol = 1
                           )
  )
)
```


### **Overriding Key `size` in Legends**

In most cases, you simply want to make the legend colors more visible for your users. Either you want to increase the size of points that are potentially too small or decrease the size of points that are just too large. Remember that aesthetic properties are inherited from data. Legend properties are inherited from the aesthetic mappings of their geoms. The legend properties may, however, benefit from modifications. When you want to change the legend properties manually, you can use the `guide()` function and specify arguments with helper functions `guide_legend()`. 

We will use `guides()` along with `guide_legend()` in order to override aesthetics. The general behavior will be to add a layer to a plot object like that shown below. 

Note: The goal of these examples is to illustrate how to change the key characteristics, not how to make everything match. 

```
guides(
     <aesthetic> = guide_legend(
                       override.aes = list(
                            <same or other aesthetic> = numeric or string value
                            )
                   )
      )
```

### **Dealing with a Constant Key Size**

When the legend provide a key that corresponds to an aesthetic other than `size`, changing the size of them does not compromise the plot integrity. We will take a single plot and adjust the size in four ways. Some points will be smaller than the default and some larger.

We will use `guides()` along with `guide_legend()` in order to override the size aesthetic using `override.aes = list(size = numeric value)`. Please note that point `size` is visibly present in default plots but not controlled by any coding. Importantly, remember that all aesthetics that you see in the plot (and some you don't see because they are invisible) are controlled in some manner, whether by you the creator or by the developers of **{ggplot2}**. In the default case, `size` is controlled but by the developers default choices.

`guides(col = guide_legend(override.aes = list(size = numeric value)))`

```{r}
suppressWarnings(
  plot(
    gridExtra::arrangeGrob(plot1 + 
                             labs(title = 'guide_legend(override.aes = list(size = 1))',
                                  tag = "A") +
                             guides(col = guide_legend(override.aes = list(size = 1))), 
                           
                           plot1 + 
                             labs(title = 'guide_legend(override.aes = list(size = 2))',
                                  tag = "B") + 
                             guides(col = guide_legend(override.aes = list(size = 2))), 
                           
                           plot1 + 
                             labs(title = 'guide_legend(override.aes = list(size = 3))',
                                  tag = "C") + 
                             guides(col = guide_legend(override.aes = list(size = 3))),
                           
                           plot1 + 
                             labs(title = 'guide_legend(override.aes = list(size = 6))',
                                  tag = "A") + 
                             guides(col = guide_legend(override.aes = list(size = 6))),
                           ncol = 2)
  )
)
```

Which override do you like best? Which is most helpful for your client? Which legend strikes the best balance between the point points and the legend points?

Of course, you really might wish to do something like reverse the legend labels as well by adding arguments.

```{r}
plot1 + 
  guides(col = guide_legend(override.aes = list(size = 3),
                            reverse = TRUE
                            )
       ) +
  labs(title = "", tag = "")
```


### **Dealing with a Variable Key Size**

If you do not like the size of the points and you are unsure what numeric value is associated with a shape, you can always control this yourself by adding a variable to the data frame and setting the `scale_size_manual()`. This way, you can adjust the legend size and ensure that the sizes in the legend correspond to the size in plot just as the default behavior works for a legend.

Here we will `mutate()` a new variable using `case_when()` that specifies a numeric value to serve as the size of points for each `Team`. We will then override the `size` of the legend keys corresponding to the `col` aesthetic by passing a two-element vector containing the same values. In the event that we want to reuse these sizes (and prevent some errors), we will assign the values to a *named* vector that we will use in both places in the plot code.

```{r}
legend_point_size <- c("Men" = 3, "Women" = 4.5)

SWIM %>%
  filter(Team != "Mixed",
         Time < 500
         ) |>
  mutate(TeamSize = case_when(
    Team == "Men" ~ legend_point_size[1],
    Team == "Women" ~ legend_point_size[2],
    )) |>
  ggplot(mapping = aes(x = Time, y = Split50)) +
  geom_point(aes(col = Team, size = TeamSize)) +
  labs(title = "default",
       tag = " --- "
       ) +
  scale_size_identity() +
  guides(col = guide_legend(override.aes = list(size = legend_point_size)))
```

You could achieve the same plot by setting numeric values specifically as with the following: 

```
guides(col = guide_legend(override.aes = list(size = c(3, 4.5))))
```


## **Overriding other aesthetics of Legend Keys**

Examples:

Change the shape, size, color, and alpha of the `col` aesthetic. Because there are **two** labels for the `col` aesthetic, we need to pass one value for a constant applied to all or a two-element vector if you wish for them to vary.

```{r}
base_plot_3 +
  guides(col  = guide_legend(override.aes = list(shape = 15,
                                                 size = 4, 
                                                 col = c("firebrick", "goldenrod"),
                                                 alpha = .3
                                                 ))
        )
```

Change the shape, size, color, fill, alpha, and stroke of the `size` aesthetic:

Because there are **three** labels for the `size` aesthetic, we need to pass three of each to vary.

```{r}
base_plot_3 +
  guides(size  = guide_legend(override.aes = list(shape = 22,
                                                  size = c(2, 4, 6), 
                                                  col = c("cornflowerblue", 
                                                          "goldenrod", 
                                                          "firebrick"
                                                          ),
                                                  fill = "grey",
                                                  alpha = .6,
                                                  stroke = 2
                                                  ))
        )
```


# **Reordering Legend Labels using `scale_<aesthetic>_manual()`**

Let's say our plot is mapping a single variable to two aesthetics, `col` and `shape`. 

```{r}
SWIM %>%
  filter(Time < 500) |>
  ggplot(mapping = aes(x = Time, y = Split50)) +
  geom_point(mapping = aes(col = Team, 
                           shape = Team
                           )
             ) 
```

We want to order the keys and labels in the legend. This is a legend that has two parts, `col` and `shape`. There are three key-label pairs to order. In general, the blue squares appear to be higher in the plot than the red circles or green triangles. The legend should at least try to match that order.  

We will need to manually adjust:

- `color` using `scale_color_manual()` 
- `shape` using `scale_shape_manual()`

Importantly, we need to know the labels names

```{r}
legend_colors <- c("Women" = "blue", "Men" = "red", "Mixed" = "gray")
legend_shapes <- c("Women" = 15, "Men" = 16, "Mixed" = 17)
legend_color_size <- 3 # make all bigger
  
SWIM %>%
  filter(Time < 500) |>
  ggplot(mapping = aes(x = Time, y = Split50)) +
  geom_point(mapping = aes(col = Team, 
                           shape = Team
                           )
             ) +
  scale_color_manual(values = legend_colors) +
  scale_shape_manual(values = legend_shapes) +
  guides(col = guide_legend(override.aes = list(size = legend_color_size)))
```


OK great, we have manually set our colors and shapes but the order is still off. The labels are not ordered in a way to facilitate perception. The named vector is not always going to work and sometimes you have to manipulation the `values`, `breaks`, and text `labels` independently.

We will:

- add labels and breaks to `scale_color_manual()` 
- add labels and breaks to `scale_shape_manual()`

The ordering of the elements in the `label` vector will control how they labels are 

For example, `labels = c("Women", "Men", "Mixed")` will reorder the text labels in that order from top to bottom. 

Let's make a `list()` that contains the values two vectors, one containing the `values` and one containing the way to place `breaks`. We will do this both for the `col` and `shape` aesthetics because they are mapped to the same variable, `Team`.

```{r}
legend_color_manual <- list(
  values = c("blue", "red", "gray"),
  breaks = c("Women", "Men", "Mixed")
)

legend_shape_manual <- list(
  values = c(15, 16, 17),
  breaks = c("Women", "Men", "Mixed")
)
```

Add the `values`:

```{r}
SWIM %>%
  filter(Time < 500) |>
  ggplot(mapping = aes(x = Time, y = Split50)) +
  geom_point(mapping = aes(col = Team, 
                           shape = Team
                           )
             ) +
  # adjust manually for col
  scale_color_manual(values = legend_color_manual$values
                     ) +
  # adjust manually for shape
  scale_shape_manual(values = legend_shape_manual$values) +
  # then override the size using the object defined earlier
  guides(col = guide_legend(override.aes = list(size = legend_color_size)))
```

Except the result is that the colors not longer match the same ordering from the named vector `legend_colors`.


Adding `breaks` to the plot, gives us:

```{r}
SWIM %>%
  filter(Time < 500) |>
  ggplot(mapping = aes(x = Time, y = Split50)) +
  geom_point(mapping = aes(col = Team, 
                           shape = Team
                           )
             ) +
  # adjust manually for col
  scale_color_manual(values = legend_color_manual$values,
                     breaks = legend_color_manual$breaks
                     ) +
  # adjust manually for shape
  scale_shape_manual(values = legend_shape_manual$values,
                     breaks = legend_shape_manual$breaks
                     ) +
  # then override the size using the object defined earlier
  guides(col = guide_legend(override.aes = list(size = legend_color_size)))
```

If you do not add the values, you will likely receive an error because a color palette needs to be passed to `values` for `scale_color_manual()`.





## **Reordering Legend Labels**

Legend labels are presented in an order, whether from top to bottom or from left to right. We have discussed previously that legend labels are not always presented alphabetically. The order depends on the variable type. When variables are character vectors, they will be ordered alphabetically but if they are factors, they will be ordered based on their order, which will differ based on them being factors or ordered factors. Using `unique()`, we can see the unique levels. 

```{r}
unique(SWIM$Team)
```

This is simply a character vector. When this type of `Team` variable is mapped to the aesthetic, the order of the labels in the legend do not map on to the spatial positioning of the data in the plot. 

Moreover, there are no levels to `Team` because character vectors don't have level. Factors have levels so when we use all`levels()` to examine the variables, we will see nothing is returned.

```{r}
levels(SWIM$Team)
```

Vectors that are factors will contain levels, so converting the vector will return its levels.

```{r}
levels(factor(SWIM$Team))
```

The levels returned make the order parent: `"Men"`, `"Mixed"`, and `"Women"`. When displayed in the legend, they will take on this order from top to bottom for the default legend orientation (e.g., `"right"`). This order will not address the the mismatch of the labels in the data. Including all three levels of the `Team` variable will make this mismatch more apparent. Such an arrangement will make cognitive processing of the visualization more challenging.

Here is plot with three levels.

```{r echo=FALSE, message=FALSE, warning=FALSE}
plot_complex_2 <- SWIM |>
   filter(Event == "Freestyle") |>
   filter(Time >= 75 & Time <= 150) |>
   ggplot(mapping = aes(x = Split50, 
                        y = Time,
                        )
          ) +
   geom_point(mapping = aes(size = Time,
                            shape = Team,
                            fill = Team,
                            col = Team
                            ),
              position = position_jitter(), 
              alpha = .7,
              col = "grey20",
              stroke = 1
              ) +
  scale_shape_manual(values = c(21, 22, 24)) +
  guides(size = "none")
```

## **Reordering Factor Levels using {forcats}**

One of the easiest ways to ensure that the order to the legend labels matches that of the data presented in the plot when you are dealing with a categorical variable is to convert the vector to a factor and reorder it based on the data. The **{forcats}** library makes this task easy using two functions, using `fct_reorder()` and  `fct_reorder2()`. 

The two functions will reorder a factor's levels by sorting them based on another variable. The main difference between the two functions is that `fct_reorder()` will reorder based on a single dimension and is thus best for 1-dimensional displays whereas `fct_reorder2()` will reorder based on two dimensions and is best for 2-dimensional displays where the factor is mapped to a non-position aesthetic. 

In order to see how the factor levels may be arranged based on the numeric variables for the scatter plot, we can use `group_by()` and `summarize()` the median, which is the default behavior of `fct_reorder()`.

```{r}
SWIM |>
  filter(Event == "Freestyle") |>
  filter(Time >= 75 & Time <= 150) |>
  group_by(Team) |>
  summarize(Time = median(Time)) |>
  ungroup() |>
  arrange(Time)
```

The means from fastest to slowest are `"Mixed"`, `"Men"`, and `"Women"`.

```{r}
SWIM |>
  filter(Event == "Freestyle") |>
  filter(Time >= 75 & Time <= 150) |>
  group_by(Team) |>
  summarize(Split50 = median(Split50)) |>
  ungroup() |>
  arrange(Split50)
```

The means for the split time at 50 m from fastest to slowest is again `"Mixed"`, `"Men"`, and `"Women"`. We need to ensure that our plot legend is from top to bottom `"Women"`, `"Men"`, and `"Mixed"` or from left to right `"Mixed"`, `"Men"`, and `"Women"`. 


### **Comparing Plots with `fct_reorder()` and `fct_reorder2()`**

You can reorder the vector in the data frame before passing to `ggplot()` or within the `aes()` mapping in the object. However, if you have multiple variable-aesthetic mappings to that variable, your more efficient approach will be to change in the data frame.

Some key features of both functions:

`.f`: the factor
`.x`: the variable for reordering with `fct_reorder()`
`.x and .y`: the variable(s) for reordering with `fct_reorder2()`


#### Using `forcats::fct_reorder()`:

Adjust the grouping of `Team` by `Split50`.

```{r}
SWIM |>
  filter(Event == "Freestyle") |>
  filter(Time >= 75 & Time <= 150) |>
  mutate(Team = forcats::fct_reorder(.f = Team, 
                                     .x = Split50
                                     )
         ) |>
  pull(Team)
```

Notice this order is `"Mixed"`, `"Men"`, and then `"Women"`. 


#### Using `forcats::fct_reorder2()`:

Adjust the grouping of `Team` by `Time` and `Split50`.

```{r}
SWIM |>
  filter(Event == "Freestyle") |>
  filter(Time >= 75 & Time <= 150) |>
  mutate(Team = forcats::fct_reorder2(.f = Team, 
                                      .x = Time, 
                                      .y = Split50
                                      )
         ) |>
  pull(Team)
```

Notice this order is `"Women"`, `"Men"`, and then `"Mixed"`. This ordering may appear odd because the mixed group is faster than men but this is outcome results from the fact that the `Distance` variable that is not accounted for in the data filtering. For illustration purposes, with our plot we don't care about that. Nevertheless, the horizontal order would be good if the legend was positioned along the top/bottom. The vertical order is problematic unless we reverse it. 


### **Plotting and Comparing Reordering using `fct_reorder()` and `fct_reorder2()`**

We will specify `.f = Team` and `.x` as `Split50` and when used, `.y = Time`.

#### Plotting with a Reordering by `forcats::fct_reorder2()`:

```{r}
SWIM |>
   filter(Event == "Freestyle") |>
   filter(Time >= 75 & Time <= 150) |>
   mutate(Team = forcats::fct_reorder2(.f = Team, 
                                       .x = Split50, 
                                       .y = Time
                                       )) |> 
   ggplot(mapping = aes(x = Split50, 
                        y = Time,
                        )
          ) +
   geom_point(mapping = aes(size = Time,
                            shape = Team,
                            fill = Team,
                            col = Team
                            ),
              position = position_jitter(), 
              alpha = .7,
              col = "grey20",
              stroke = 1
              ) +
  scale_shape_manual(values = c(21, 22, 24)) +
  guides(size = "none")
```

When the legend is positioned to the right of the plot, the vertical positioning of the legend labels now matches the data.

To position the legend at the bottom of the plot, we get:

```{r}
SWIM |>
   filter(Event == "Freestyle") |>
   filter(Time >= 75 & Time <= 150) |>
   mutate(Team = forcats::fct_reorder2(.f = Team, 
                                       .x = Split50, 
                                       .y = Time
                                       )) |> 
   ggplot(mapping = aes(x = Split50, 
                        y = Time,
                        )
          ) +
   geom_point(mapping = aes(size = Time,
                            shape = Team,
                            fill = Team,
                            col = Team
                            ),
              position = position_jitter(), 
              alpha = .7,
              col = "grey20",
              stroke = 1
              ) +
  scale_shape_manual(values = c(21, 22, 24)) +
  guides(size = "none") +
  theme(legend.position = "bottom")
```

When the legend is positioned at the bottom, the horizontal positioning of the legend labels does not match the data. You can also change the `.x` and `.y` variables if necessary.


### **Plotting with a Reordering by `forcats::fct_reorder()`**:

`fct_reorder()` will reorder `Team` only by a single variable. You could choose either `Split50` or `Time`.

#### Bar plots

When you have a bar plot, reordering the factor will help arrange the data from lowest to highest, thus making the data more easy to perceive. When dealing with variables plotting a continuous and a discrete variable, use `fct_reorder()`.

```{r}
SWIM |>
   filter(Event == "Freestyle") |>
   filter(Time >= 75 & Time <= 150) |>
   mutate(Team = forcats::fct_reorder(.f = Team, .x = Time)) |> 
   ggplot(mapping = aes(x = Team, 
                        y = Time,
                        )
          ) +
   geom_boxplot(mapping = aes(fill = Team))
```

Although we have ordered the box plots, the legend is not in an order that matches the vertical ordering. If you want the legend labels oriented vertically, consider adjusting them using `labels` and `breaks` settings with `scale_*_manual()` functions. However, moving the legend to the bottom, top, or changing the `direction` to horizontal would suffice. You can also consider direct labeling of the plot.

```{r}
SWIM |>
   filter(Event == "Freestyle") |>
   filter(Time >= 75 & Time <= 150) |>
   mutate(Team = forcats::fct_reorder(.f = Team, .x = Time)) |> 
   ggplot(mapping = aes(x = Team, 
                        y = Time,
                        )
          ) +
   geom_boxplot(mapping = aes(fill = Team)) +
   theme(legend.position = "bottom")
```


## **Reordering the Labels in a Legend**

This topics relates to creating perceptually-efficient data visualizations nevertheless we will address this topic. The legend information is supposed to support the data presented in a plot. Sometimes, however, the ordering of the labels in the legend compromises perceptual processing by the user. 

Let's take another look at a plot. We will remove the `size` legend just to reduce confusion. Remember we can do this using `guides(<aesthetic> = "none")`. 

```{r}
base_plot_3 + 
  guides(size = "none")
```

In this plot, the legend label order is opposite that of the data. Men are faster than Women so processing the `fill` and the `shape` aesthetics but the legend is arranged in the reversed order. There is no utility in ordering the labels in a way that increases the cognitive demand on the user. Not paying attention to such issues may result in your plots being less effective than is necessary. Although there are often [desirable difficulties](https://bjorklab.psych.ucla.edu/wp-content/uploads/sites/13/2016/04/EBjork_RBjork_2011.pdf) associated with increased cognitive effort, the trade off here is a misinterpretation of the plot.

What can we do? Well, we already discussed changing the legend position by modifying `legend.position`. We can move the legend to the bottom (below the plot). By doing so, the left-right arrangement matches the location of the data along the x axis.

```{r}
base_plot_3 + 
  guides(size = "none") +
  theme(legend.position = "bottom") 
```

But let's say either we do not want to position a legend along the top or bottom or that doing so does not solve the problem. The more levels and labels there are, the more difficult this will be do achieve. We will need to rearrange the labels themselves.

```{r}
base_plot_3 + 
  guides(size = "none") +
  theme(legend.position = "right") 
```

```{r include=FALSE}
#guides(alpha = guide_legend(title = "Transparency"))
```


# **Changing Title Characteristics**

If your mapped variable is messy, you will need to make it look pretty. Perhaps it contains an under_score, is all lowercase, or is CamelCase, etc. You can change the title. Because the legend corresponds to aesthetics added to the plot beyond the x and y variables, we need to specify them individually. This means that you have control over each legend. 

Changing the title for each legend is easy using `guide_legend(title = "")`. Just remember that if a variable is mapped to multiple aesthetics, you will need to change the title in both places or the grouped legend will be split into its parts.

Example:

````
guides(
  col = guide_legend()
  shape = guide_legend()
  size = guide_legend()
)
```

## **Changing Legend Title `guides()`**

```{r}
suppressMessages(
  plot(
    gridExtra::arrangeGrob(base_plot_3 + 
                             labs(title = 'default'),
                           
                           # fill only
                           base_plot_3 + 
                             guides(fill = guide_legend(title = "Teams")) +
                             labs(title = 'change fill only'),
                           
                           # fill, shape, col
                           base_plot_3 + 
                             guides(fill = guide_legend(title = "Teams"),
                                    shape = guide_legend(title = "Teams"),
                                    col = guide_legend(title = "Teams"),
                                    ) +
                             labs(title = 'change col, fill, and shape'),
                           
                           ncol = 1
    ))
  )
```

## **Changing Legend Title Position**

You don't need to change the title using `labs()`. Here, we change the `title` and `title.position` for the `col` aesthetic only within `guide_legend()`. Adding other aesthetic changes would be as simple as specifying them in `guides()`.

```{r}
base_plot_3 + guides(col = guide_legend(title = "New Title", 
                                        title.position = "left"
                                       )
                   )

```

## **Changing Legend Direction and Label Position**

Here, we also change the `direction` and `label.position` for the `col`, `size`, and `shape` in `guides()`.

```{r}
plot3 + guides(
  # color aesthetic  
  col = guide_legend(title = "Color Title",
                     direction = "horizontal",
                     title.position = "bottom",
                     label.position = "top"
                     ),
  # the size dimension
  size = guide_legend(title = "Size Title",
                      direction = "vertical",
                      title.position = "top",
                      label.position = "top"
                      ),
    # the shape aesthetic (does not appear because point all all the same shape) 
    shape = guide_legend("Shape Title")
 ) + 
  theme(legend.position = "bottom")
```

Whether this outcome is appropriate is up for discussion. When you need to change such legend elements, however, the above examples will be helpful.


# **Session Info**

```{r, echo = FALSE}
sessionInfo()
```

