---
title: "**Summarizing Data: Using `across()` and Lists**"
author: "Gabriel I. Cook"
#date: "`r Sys.Date()`"
#date: "`r format(Sys.time(), '%d %B, %Y')`"

execute:
  #enabled: false
  freeze: auto
---

```{r message=FALSE, warning=FALSE, include=FALSE}
# secret functions
# source(here::here("r", "src", "color_format_text.R"))
```

```{r}
#| label: load-packages
#| include: false

#source(here::here("r", "my_functions.R"))
#source(here::here("r", "functions", "view_html.R"))
R.utils::sourceDirectory(here::here("r", "functions"))
```


# **Overview**

This module introduces some somewhat *advanced* approaches to summarizing data. They are advanced insofar as they involve creating lists of summary functions and using `dplyr::across()` to run those functions *across* multiple variables. They are more complicated than adding lines of expressions to `summarize()` but the cost associated with greater complexity buy you simplicity of code and reproducibility. 

# **Libraries**

```{r message=FALSE, warning=FALSE, include=TRUE}
library(tibble)
library(dplyr)
```

# **Create a Data Frame**

```{r}
DATA <- data.frame(
  id = 1:6,
  group = c(rep("a", 3), rep("b", 3)),
  x = c(11, 21, 13, 51, 44, NA),
  y = c(30, 20, 33, NA, 20, 19)
  )

DATA <- tribble(
  ~id, ~group, ~x, ~y,
  1, "a", 11, 30,
  2, "a", 21, 20,
  3, "a", 13, 33,
  4, "b", 51, NA,
  5, "b", 44, 20,
  6, "b", NA, 19
)
```

# **Standard Data-Summary Techniques**

## **Summarizing Vectors**

```{r}
mean(DATA$x, na.rm = T)
mean(DATA$y, na.rm = T)
median(DATA$x, na.rm = T)
median(DATA$y, na.rm = T)
```

**NOTE:** The mean and median of `x` **differ**.

## **Summarizing Vectors in Data Frames**

### **1 Variable, 2 Metrics**

```{r}
DATA |>
  summarize(x = mean(x, na.rm = TRUE),
            x_median = median(x, na.rm = TRUE)
            )
```

**Warning:** The mean and median of `x` are **the same**. The `median()` is computed based on the *new* value of `x` that is assigned by the first line in `summarize()` (e.g `x = mean(x, na.rm = TRUE)`). You would want to use a *new* variable name.

Solution: Assigning to names other than `x`:

```{r}
DATA |>
  summarize(mean = mean(x, na.rm = TRUE),
            median = median(x, na.rm = TRUE)
            )
```


### **2 Variables, 1 Metric**

```{r}
DATA |>
  summarize(x_mean = mean(x, na.rm = TRUE),
            y_mean = mean(y, na.rm = TRUE),
            )

```

**NOTE:** Are the means really the same?

### **2 Variables, 2 Metrics**

```{r}
DATA |>
  summarize(x_mean = mean(x, na.rm = TRUE),
            y_mean = mean(y, na.rm = TRUE),
            x_median = median(x, na.rm = TRUE),
            y_median = median(y, na.rm = TRUE)
            )

```

### **2 Variables, 2 Metrics with Grouping**

```{r}
DATA |>
  group_by(group) |>
  summarize(x_mean = mean(x, na.rm = TRUE),
            y_mean = mean(y, na.rm = TRUE),
            x_median = median(x, na.rm = TRUE),
            y_median = median(y, na.rm = TRUE)
            )
```


# **Advanced Data-Summary Techniques**

Coding each variable to include in the summarized data frame can be tedious. 

- use `across()`
- pass a list of functions

## **Summarizing Across with `dplyr::across()`**

`across()` is used when you want to iterate a function or set of functions across a multiple variables. The function will require you to pass arguments for the columns you want to summarize, the function(s) specifying how to summarize, and the names of the new output variables. By default, the 

```{.r}
across(.cols, 
       .fns, ..., 
       .names = NULL, 
       .unpack = FALSE
       )
```

## **`dplyr::across()`: Parameters/Arguments**

- `.cols`: the columns to perform a function upon 
- `.fns`: the function(s) to apply to the column in `.cols`
- `.names`: a glue specification that describes how to name the output columns; use `{.col}` to stand for the selected column name, and `{.fn}` for the function being applied; defaults to `"{col}_{fn}"`


## **`dplyr::across()`: Passing Arguments**

- `.cols = c(x, y)`
- `.fns = ~mean(x, na.rm = TRUE)`

Accepting the default argument for `.names`:

- `.names = NULL`

```{r}
DATA |>
  group_by(group) |>
  summarize(across(.cols = c(x, y), 
                   .fns = ~mean(.x, na.rm = TRUE)
                   )
            )
```

Passing an argument to `.names`:

- `.names = "{col}_{fn}"`

```{r}
DATA |>
  group_by(group) |>
  summarize(across(.cols = c(x, y), 
                   .fns = ~mean(.x, na.rm = TRUE),
                   .names = "{col}_{fn}"
                   )
            )
```

Or using a quoted vector: `.cols = c("x", "y")`

```{r}
DATA |>
  group_by(group) |>
  summarize(across(.cols = c("x", "y"), 
                   .fns = ~mean(.x, na.rm = TRUE),
                   .names = "{col}_{fn}"
                   )
            )
```

Or passing a quoted vector: 

- must use `all_of()` or `any_of()` for variable selection
- `.cols = all_of(summarize_these)`
- `.cols = summarize_these` will produce a warning

```{r}
summarize_these <- c("x", "y")  # create the vector to pass

DATA |>
  group_by(group) |>
  summarize(across(.cols = any_of(summarize_these), 
                   .fns = ~mean(.x, na.rm = TRUE),
                   .names = "{col}_{fn}"
                   )
            )
```

**Note:** defining a vector of variables can be a helpful solution when you have multiple summary tables for which you use the same variables.  

## **Annoyances with `across()`**

- More complicated
- Although {col} is useful (e.g., `x` and `y`), {fn} results in a numeric value which is not diagnostic of the function



# **Using Lists**

- Lists hold all different types of objects as their elements
- Vectors, data frames, functions

```{r}
my_list <- 
  list(
    num_vect = c(1, 2),
    char_vect = c("2", "3", "5"),
    dataframe = DATA,
    afunction = function(x) {mean(x)}
    )
```

# **Silly Example**

- Call the function element to
- Perform on the numeric vector element

```{r}
my_list$afunction(my_list$num_vect)
```

## **Passing a List of Functions to `.fns`**

- `.cols = c(x, y)`
- `.fns = list(~mean(x, na.rm = TRUE))`

```{r}
DATA |>
  group_by(group) |>
  summarize(across(.cols = c(x, y), 
                   .fns = list(~mean(.x, na.rm = TRUE)),
                   .names = "{col}_{fn}"
                   )
            )
```

**Note:** The `~` is use as a lambda-like operator that results in iterating the function over all instances of *x*. In this case, `list(~mean(x, na.rm = TRUE)`, the *x* is not referring to the `x` column in the data frame but instead the values in all variables passed to `.cols`. In this case, the *x* would be both `x` and `y`, in that order. 


## **Same Annoyances**

- More complicated (requires remembering the function, in the list, and the ~)
- Although {col} is useful (e.g., `x` and `y`), {fn} results in a numeric value which is not diagnostic of the function


## **Giving the List Elements Names**

- `.cols = c(x, y)`
- `.fns = list(some_name = ~mean(x, na.rm = TRUE))`

```{r}
DATA |>
  group_by(group) |>
  summarize(across(.cols = c(x, y), 
                   .fns = list(some_name = ~mean(.x, na.rm = TRUE)),
                   .names = "{col}_{fn}"
                   )
            )
```

## **Annoyances**

- Still complicated (requires remembering the function, in the list, and the ~)


## **Passing a List Object to `.fns`**

- create `summary_funcs` list of function(s)
- `.fns = summary_funcs`

Create a list containing `~mean(x, na.rm = TRUE))` used in previous example:

```{r}
summary_funcs <- list(
  mean = ~mean(na.omit(.x))  
  )
```

Then pass to `.fns`, `.fns = summary_funcs`:

```{r}
DATA |>
  summarize(across(.cols = c(x, y), 
                   .fns = summary_funcs,
                   .names = "{col}_{fn}"
                   )
            ) 
```

Pair with `.cols = summarize_these` to summarize the variables in `summarize_these` using the function(s) in `summary_funcs`: 

```{.r}
DATA |>
  summarize(across(.cols = summarize_these,
                   .fns = summary_funcs,
                   .names = "{col}_{fn}"
                   )
            ) 
```

**Note:** If you try to execute this, pay attention to the warning Message!

```
! Using an external vector in selections was deprecated in tidyselect 1.1.0.
ℹ Please use `all_of()` or `any_of()` instead.
  # Was:
  data %>% select(summarize_these)

  # Now:
  data %>% select(all_of(summarize_these))
  
See <https://tidyselect.r-lib.org/reference/faq-external-vector.html>.  
```

**Solution:** When passing an external vector, you need to use `all_of()` or `any_of()`.

```{r}
DATA |>
  summarize(across(.cols = all_of(summarize_these),
                   .fns = summary_funcs,
                   .names = "{col}_{fn}"
                   )
            ) 
```



## **Annoyances/Benefits**

- Requires creating other objects
- Simplifies the code
- Does not require remembering the functions, the list, and ~ after created once 


## **Adding More Functions as List Elements**

Add functions to the list to accomplish more

```{r}
summary_funcs <- list(
  mean = ~mean(na.omit(.x)),
  median = ~median(na.omit(.x)),
  sd = ~sd(na.omit(.x)),
  n = ~length(na.omit(.x))
  )
```


## **Passing the List Object to `.fns`: Grouping**

- `group_by(group)`    
- `.fns = summary_funcs`

```{r}
DATA |>
  group_by(group) |>
  summarize(across(.cols = all_of(summarize_these),
                   .fns = summary_funcs,
                   .names = "{col}_{fn}"
                   )
            )
```

## **Annoyances/Benefits**

- Creating and remembering the list object name
- **Solution:** Add an .R script to your functions directory and make a code snippet for the object name. 
