---
title: "Class Exercise: Read, Manipulate, and Summarize Data"
date: "`r Sys.Date()`"
output:
  html_document:
    always_allow_html: yes
    toc: yes
    number_sections: yes
    code_folding: hide
  pdf_document:
    toc: yes
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



*Enter your name here*   


# **Load relevant libraries**

Load **{readr}**, **{magrittr}**, **{dplyr}**, and **{lubridate}**.

```{r}
library("magrittr")
library(readr)
library(dplyr)
library(lubridate)
```


# **Read a `.csv` file from a URL**

The following data are taken from [tf.tfrrs.org](https://tf.tfrrs.org/all_performances/CA_college_m_Claremont_Mudd_Scripps.html?list_hnd=600&season_hnd=131).

The file referenced at the url below contains data from a particular year and for men on the CMS track & field team who participated in an event during different meets. 

url: https://github.com/slicesofdata/dataviz23/raw/main/data/tfrrs/out_2010_CMS_100.csv

Read the this file and assign it to an object holding the data frame.

```{r}
source(here::here("r", "my_functions.R"))

DAT <- readr::read_csv("https://github.com/slicesofdata/dataviz23/raw/main/data/tfrrs/out_2010_CMS_100.csv")

```


# **Take a glimpse at the structure of the data frame**

```{r}
glimpse(DAT)
```


What are the names and types variables in the data frame?

Consider whether you understand the variables and what the values represent. If not, make a note because you may wish to ask the CMS T&F liaison these questions during a meeting. 



# **Mutating Variables**

Create a new variable to represent the year those meets took place (see file name) and another variable that represents the event type (see file name). You may also wish to create a variable for whether the athlete is either a Stag or Athena.

```{r}
DAT <- DAT %>% 
  dplyr::mutate(.,
    Class = Year,
    Year = 2010
    ) 

DAT %>%
  view_html()

```


# **Summarize Data**

Summarize the data in at least two ways.

```{r}
DAT %>%
  group_by(., Athlete) %>%
  summarize(., 
            Mean = mean(na.omit(Time))
  )


```


```{r}



```

Did **{dplyr}** throw an error for the `summarize()` function? 



# **Summarize your finding**

Describe something that you found based on your summary.





# **Problem x: Write the modified data file**

Write the new file to your local disk, specifically at `/data` using the **{here}** library.

Do not save it as the same name.

```{r}
write_csv(x = DAT,
          file = here::here("data", "cleaned_out_2010_CMS_100.csv")
)
```




**Please knit an `html` file.** 
