---
title: "Homework 08"
author: "your-name"
#date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: false
    number_sections: true
   # code_folding: show
  pdf_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


For this homework, used this `R Markdown` (e.g., `.Rmd`) file and save it to your `/report` project directory on your computer with the name and a prefix `hw08-Lastname-FirstInitial`. Full name example: `hw08cookg.Rmd`. Make sure to add your name to the YAML code above.


You will take a plot that you are working on for your project and modify it in two ways based on the legends and arrangement module. 

Load your data in the following code block.

```{r message=FALSE, warning=FALSE, results='hide'}



```

1. Create your base plot object.

```{r fig.align='center', message=FALSE, warning=FALSE, out.width="60%"}



```

2. Modify your base plot to implement your first adjustment.

- *Describe why you might want to implement this approach*



- *Add your code to the base plot in the code block below*

```{r fig.align='center', message=FALSE, warning=FALSE, out.width="60%"}




```

2. Modify your base plot to implement your second adjustment.

- *Describe why you might want to implement this approach*





- *Add your code to the base plot in the code block below*

```{r message=FALSE, warning=FALSE, fig.align='center',out.width="60%"}




```


Finally, knit an `html` file and upload to:
https://ln5.sync.com/dl/a038628f0/wwfifjxk-f7rfshin-rkedi3y8-77f9zaii
