---
title: 'Exercise: Linear Models'
output:
  pdf_document:
    toc: false
  html_document:
    toc: false
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
eval_answers <- TRUE
include_answers <- FALSE
```

```{r, include=FALSE}
# source(here::here("r", "my_functions.R"))                # if your functions are in a file
R.utils::sourceDirectory(here::here("r", "functions"))
```

# **Class Practice**

You can practice utilizing linear models to answer question about your project data. 

# **Loading Libraries**

Load libraries that you will need to interpret a linear model. Add to the list as you realize you need them.

```{r, eval=eval_answers, include=include_answers}
#library(ggplot2)
```

# **Reading Data**

Obtain project that for running a linear model. 

```{r, eval=eval_answers, include=include_answers}
#GNG <- readRDS(here::here("data", "gng", "gng_long_clean.Rds"))
```

```{r, eval=eval_answers, include=include_answers}

```

# **Fitting the Model**

Fit your data to a linear model for which you predict a single outcome variable using one or more predictor variables.



# **Interpreting the Model Fit**

Interpret whether the overall fit of the model by including the metrics relevant for arriving at your conclusion. 



# **Interpreting Model Coefficients**

Provide the model coefficients and interpret those coefficients accordingly.


# **Visualizing the Relationship(s)**

Consider ways that you might visually convey the relationships captured by the model. 




